---
title: "Norwegian Electricity & Hydro Power - Exploratory Analysis"
author: "Carla de Erausquin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(fpp3)
library(readr)
library(tidyr)
library(dplyr)
library(stringr)
library(lubridate)
library(tseries)
library(strucchange)
library(knitr) 
```

## 1. Load and Prepare the Data
```{r load-data}
# Read the raw file first to see exact column names
raw <- read_delim("data/hydro-dataset.csv", 
                  delim = ";", 
                  skip = 1,
                  col_names = TRUE)

# Check the actual column names
colnames(raw)[1:5]
nrow(raw)
```

```{r wrangle}
# Rename the first column explicitly using its position
# and then pivot

long <- raw %>%
  # Rename first column by position to avoid special character issues
  rename_with(~ "series", .cols = 1) %>%
  # Remove any completely empty rows
  filter(!is.na(series), series != "") %>%
  pivot_longer(
    cols      = -series,
    names_to  = "month_raw",
    values_to = "MWh"
  ) %>%
  mutate(
    # Extract "1993M01" pattern from column name
    month_code = str_extract(month_raw, "[0-9]{4}M[0-9]{2}"),
    year_part  = as.integer(substr(month_code, 1, 4)),
    mon_part   = as.integer(substr(month_code, 6, 7)),
    Month      = yearmonth(make_date(year_part, mon_part, 1)),
    MWh        = as.numeric(MWh)
  ) %>%
  filter(!is.na(Month), !is.na(MWh)) %>%
  select(series, Month, MWh)

# Verify
print(long, n = 10)
cat("\nUnique series:\n")
print(unique(long$series))
cat("\nDate range:", 
    as.character(min(long$Month)), "to", 
    as.character(max(long$Month)), "\n")
```

```{r split-tsibbles}
total_elec <- long %>%
  filter(grepl("Total", series)) %>%
  select(Month, MWh) %>%
  as_tsibble(index = Month)

hydro <- long %>%
  filter(grepl("Hydro", series)) %>%
  select(Month, MWh) %>%
  as_tsibble(index = Month)

total_elec
hydro
```


---

## 2. Time Plots

```{r timeplot-total}
total_elec %>%
  autoplot(MWh) +
  labs(
    title = "Total Electricity Generation in Norway",
    subtitle = "May 1993 – March 2026",
    y = "MWh",
    x = NULL
  )
```

```{r timeplot-hydro}
hydro %>%
  autoplot(MWh) +
  labs(
    title = "Hydro Power Generation in Norway",
    subtitle = "May 1993 – March 2026",
    y = "MWh",
    x = NULL
  )
```

```{r timeplot-both}
# Plot both series together for comparison
long %>%
  mutate(Month = yearmonth(Month)) %>%
  as_tsibble(index = Month, key = series) %>%
  autoplot(MWh) +
  labs(
    title = "Norway: Total vs Hydro Electricity Generation",
    y = "MWh", x = NULL
  ) +
  theme(legend.position = "bottom")
```

> **What do we see?**  
> - Strong seasonal pattern in both series (high in winter, low in summer — Norway's hydro reservoirs fill in spring/summer but demand peaks in cold months).  
> - No obvious long-run trend; production fluctuates around a relatively stable level.  
> - Hydro tracks Total very closely, confirming hydro dominates Norwegian electricity generation.

---

## 3. Seasonal Plots

```{r seasonal-total}
total_elec %>%
  gg_season(MWh, labels = "both") +
  labs(
    title = "Seasonal Plot: Total Electricity Generation",
    y = "MWh"
  )
```

```{r seasonal-hydro}
hydro %>%
  gg_season(MWh, labels = "both") +
  labs(
    title = "Seasonal Plot: Hydro Power Generation",
    y = "MWh"
  )
```

> **Interpretation:**  
> - Generation is highest in **January/December** and lowest in **June/July**.  
> - The shape is consistent across years, confirming stable seasonality.  
> - Some years (dry years) show notably lower summer generation.

---

## 4. Subseries Plots

```{r subseries-total}
total_elec %>%
  gg_subseries(MWh) +
  labs(
    title = "Subseries Plot: Total Electricity Generation",
    y = "MWh"
  )
```

```{r subseries-hydro}
hydro %>%
  gg_subseries(MWh) +
  labs(
    title = "Subseries Plot: Hydro Power Generation",
    y = "MWh"
  )
```

> **Interpretation:**  
> - Blue horizontal lines show each month's mean — clearly higher in winter months.  
> - Within-month trends appear relatively flat, suggesting no strong long-term directional change in any particular month.

---

## 5. Lag Plots

```{r lag-hydro}
hydro %>%
  gg_lag(MWh, geom = "point", lags = 1:12) +
  labs(title = "Lag Plot: Hydro Power Generation")
```

> **Interpretation:**  
> - Strong positive correlation at **lag 12**, confirming annual seasonality.  
> - Lag 6 shows weaker/negative correlation — opposite season.

---

## 6. Autocorrelation Function (ACF)


```{r acf-pacf-total}
total_elec %>%
  gg_tsdisplay(MWh, plot_type = "partial", lag_max = 48) +
  labs(title = "ACF & PACF: Total Electricity Generation")
```

```{r acf-pacf-hydro}
hydro %>%
  gg_tsdisplay(MWh, plot_type = "partial", lag_max = 48) +
  labs(title = "ACF & PACF: Hydro Power Generation")
```

**Interpretation:**  
The ACF and PACF plots reveal the core structure of Norwegian hydro power generation. The ACF shows large positive spikes at lags 12, 24, 36 and 48, and large negative spikes at lags 6, 18 and 30, confirming a very strong and stable annual seasonal cycle where winter months are consistently high and summer months consistently low. The slow decay of these seasonal spikes across many years indicates the series has a long seasonal memory. The PACF however tells a more precise story — after controlling for intermediate months, only lag 1 (last month), lags 2–3 (with a negative direct effect), and lag 12 (same month last year) show meaningful direct relationships. The spikes at lags 24 and 36 visible in the ACF essentially disappear in the PACF, revealing they were indirect chain effects flowing through lag 12 rather than genuine direct relationships. Together, these plots suggest that a seasonal ARIMA model with an autoregressive component at both the monthly and annual level — written as SARIMA(1,0,0)(1,0,0)[12] — would be a strong starting point for modelling and forecasting Norwegian hydro power generation.



```{r stationarity-tests}
# --- ADF Test ---
# H0: non-stationary (has unit root)
# p < 0.05 → reject H0 → stationary
adf_hydro <- adf.test(hydro$MWh)
adf_total <- adf.test(total_elec$MWh)

cat("=== ADF Test ===\n")
cat("Hydro:  p-value =", round(adf_hydro$p.value, 4), "\n")
cat("Total:  p-value =", round(adf_total$p.value, 4), "\n")

# --- KPSS Test ---
# H0: stationary
# p < 0.05 → reject H0 → non-stationary
kpss_hydro <- hydro %>% features(MWh, unitroot_kpss)
kpss_total <- total_elec %>% features(MWh, unitroot_kpss)

cat("\n=== KPSS Test ===\n")
print(kpss_hydro)
print(kpss_total)

# --- How many differences needed? ---
ndiff_hydro <- hydro %>% features(MWh, unitroot_ndiffs)
nsdiff_hydro <- hydro %>% features(MWh, unitroot_nsdiffs)

cat("\n=== Differences Needed ===\n")
cat("Regular differences:  ", ndiff_hydro$ndiffs, "\n")
cat("Seasonal differences: ", nsdiff_hydro$nsdiffs, "\n")
```
The ADF and KPSS stationarity tests produced conflicting results for both series, with the ADF rejecting non-stationarity (p = 0.01) while the KPSS simultaneously rejected stationarity (p = 0.01). This apparent contradiction is best explained by the strong seasonal pattern already visible in the ACF — the series does not contain a permanent trend or random walk, satisfying the ADF, but its persistent annual cycle creates a form of seasonal non-stationarity that the KPSS detects. This is confirmed by the automatic differencing tests, which recommend applying both one regular difference and one seasonal difference before modelling.
---

## 7. Hydro as a Share of Total Generation

```{r hydro-share}
share <- inner_join(
  total_elec %>% rename(total = MWh),
  hydro       %>% rename(hydro  = MWh),
  by = "Month"
) %>%
  mutate(hydro_share = hydro / total * 100)

share %>%
  autoplot(hydro_share) +
  labs(
    title = "Hydro as % of Total Electricity Generation — Norway",
    y = "Share (%)", x = NULL
  ) +
  geom_hline(yintercept = mean(share$hydro_share), linetype = "dashed", colour = "steelblue") +
  annotate("text", x = min(share$Month), 
           y = mean(share$hydro_share) + 0.5,
           label = paste0("Mean: ", round(mean(share$hydro_share), 1), "%"),
           hjust = 0, colour = "steelblue")
```

> Norway's hydro share hovers around **90–95%** on average, with dips in dry years when thermal/wind backup is needed.

---

## 8. Transformation Check

```{r transformation}
# Check if a log or Box-Cox transformation stabilises variance
lambda <- hydro %>%
  features(MWh, features = guerrero)
lambda

hydro %>%
  autoplot(box_cox(MWh, lambda$lambda_guerrero)) +
  labs(
    title = paste0("Hydro Generation — Box-Cox (λ = ",
                   round(lambda$lambda_guerrero, 3), ")"),
    y = "Transformed MWh"
  )
```

λ is close to 1, little transformation is needed — the variance is already stable.  


---

## 9. STL Decomposition

```{r stl-total}
total_elec %>%
  model(STL(MWh ~ trend(window = 13) + season(window = "periodic"))) %>%
  components() %>%
  autoplot() +
  labs(title = "STL Decomposition: Total Electricity Generation")
```

```{r stl-hydro}
hydro %>%
  model(STL(MWh ~ trend(window = 13) + season(window = "periodic"))) %>%
  components() %>%
  autoplot() +
  labs(title = "STL Decomposition: Hydro Power Generation")
```

> **Interpretation:**  
> - The **trend** component is relatively flat with some multi-year cycles (wet/dry hydro years).  
> - The **seasonal** component is large relative to the series, dominating variation.  
> - The **remainder** captures unusual years — look for spikes corresponding to extreme weather events or droughts.

---

## 10. Summary Statistics

```{r summary}
# Annual totals
hydro %>%
  mutate(Year = year(Month)) %>%
  as_tibble() %>%
  group_by(Year) %>%
  summarise(
    Annual_GWh = sum(MWh, na.rm = TRUE) / 1000,
    .groups = "drop"
  ) %>%
  ggplot(aes(x = Year, y = Annual_GWh)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Annual Hydro Power Generation in Norway",
    y = "GWh", x = NULL
  )
```

## 11. Structural Break Analysis

A structural break occurs when the statistical properties of a time series
change permanently at some point in time — such as a shift in the mean level
of hydro production due to drought, new capacity, or energy policy changes.
We follow the same approach as in the lecture: visual inspection, identifying
the largest shift, the Chow test, and the automated Sup-F (QLR) test.


### 11.1 Identifying the Largest Shift

We use the `shift_level_max` feature to identify the period with the
largest shift in the mean level of the series. This feature computes
the maximum difference between the means of two adjacent windows of
data (default window size = 5 periods).

```{r shift-level-hydro}
# Find the largest level shift in hydro generation
shift_hydro <- hydro %>%
  features(MWh, features = list(shift_level_max))

print(shift_hydro)

# Identify which month this corresponds to
cat("\nThe largest shift occurs at observation:", 
    shift_hydro$shift_level_index, "\n")

hydro %>%
  slice(shift_hydro$shift_level_index) %>%
  kable(digits = 2, align = "c",
        caption = "Month with the Largest Level Shift — Hydro")
```

```{r shift-level-total}
# Find the largest level shift in total electricity
shift_total <- total_elec %>%
  features(MWh, features = list(shift_level_max))

print(shift_total)

cat("\nThe largest shift occurs at observation:", 
    shift_total$shift_level_index, "\n")

total_elec %>%
  slice(shift_total$shift_level_index) %>%
  kable(digits = 2, align = "c",
        caption = "Month with the Largest Level Shift — Total Electricity")
```

> **Interpretation:** The `shift_level_max` value represents the maximum
> difference between the means of two adjacent windows. A larger value
> means a more dramatic shift in the level of the series at that point.

---

### 11.2 The Chow Test — Is the Shift Statistically Significant?

The Chow test compares two linear models fit **before and after** the
detected break point. If there is a statistically significant difference
between the two models, this is evidence for a structural break.

- **H₀:** There is no structural break
- **H₁:** There is a structural break at this point

```{r chow-hydro}
# Prepare data with a lagged variable for the regression
hydro_df <- hydro %>%
  as_tibble() %>%
  mutate(Lag1 = lag(MWh)) %>%
  filter(!is.na(Lag1))

# Define the regression formula
# Is this month explained by last month?
reg_formula_hydro <- MWh ~ Lag1

# Run Chow test at the detected break point
chow_hydro <- sctest(
  reg_formula_hydro,
  data  = hydro_df,
  type  = "Chow",
  point = shift_hydro$shift_level_index
)

cat("=== Chow Test: Hydro Power Generation ===\n")
print(chow_hydro)

cat("\nInterpretation:\n")
if (chow_hydro$p.value < 0.05) {
  cat("p <  0.05 → Reject H0 → Structural break IS statistically significant\n")
} else {
  cat("p >= 0.05 → Fail to reject H0 → No significant structural break\n")
}
```

```{r chow-total}
# Prepare total electricity data
total_df <- total_elec %>%
  as_tibble() %>%
  mutate(Lag1 = lag(MWh)) %>%
  filter(!is.na(Lag1))

reg_formula_total <- MWh ~ Lag1

chow_total <- sctest(
  reg_formula_total,
  data  = total_df,
  type  = "Chow",
  point = shift_total$shift_level_index
)

cat("=== Chow Test: Total Electricity Generation ===\n")
print(chow_total)

cat("\nInterpretation:\n")
if (chow_total$p.value < 0.05) {
  cat("p <  0.05 → Reject H0 → Structural break IS statistically significant\n")
} else {
  cat("p >= 0.05 → Fail to reject H0 → No significant structural break\n")
}
```

---

### 11.3 The Automated Sup-F (QLR) Test

In practice we often do not know the exact date of a break. The
**Quandt Likelihood Ratio (QLR)** test — also called the Sup-F test —
runs a Chow test at **every possible point** in the series and takes
the maximum F-statistic. This identifies both whether a break exists
and where it most likely occurred.

```{r qlr-hydro}
# Calculate F-statistics at every possible break point
# from = 0.15 trims the first and last 15% to ensure enough
# observations on both sides of each candidate break point
qlr_hydro <- Fstats(
  reg_formula_hydro,
  data = hydro_df,
  from = 0.15
)

# Formal Sup-F test
supf_hydro <- sctest(qlr_hydro, type = "supF")
cat("=== Sup-F Test: Hydro Power Generation ===\n")
print(supf_hydro)

# Extract the optimal breakpoint
bp_hydro_auto <- breakpoints(qlr_hydro)

cat("\nOptimal breakpoint at observation:", bp_hydro_auto$breakpoints, "\n")

hydro_df %>%
  slice(bp_hydro_auto$breakpoints) %>%
  kable(digits = 2, align = "c",
        caption = "Optimal Breakpoint — Hydro Power Generation")
```

```{r qlr-total}
qlr_total <- Fstats(
  reg_formula_total,
  data = total_df,
  from = 0.15
)

supf_total <- sctest(qlr_total, type = "supF")
cat("=== Sup-F Test: Total Electricity Generation ===\n")
print(supf_total)

bp_total_auto <- breakpoints(qlr_total)

cat("\nOptimal breakpoint at observation:", bp_total_auto$breakpoints, "\n")

total_df %>%
  slice(bp_total_auto$breakpoints) %>%
  kable(digits = 2, align = "c",
        caption = "Optimal Breakpoint — Total Electricity Generation")
```

---

### 11.4 Visualising the F-Statistics

The F-statistics plot shows the test statistic at every possible break
point. The **red line** is the critical value — points above it indicate
statistically significant breaks. The **peak** is the most likely break point.

```{r fstats-plot-hydro}
plot(qlr_hydro,
     main = "Sup-F Test: Searching for Breakpoint — Hydro Generation")
lines(breakpoints(qlr_hydro))
```

```{r fstats-plot-total}
plot(qlr_total,
     main = "Sup-F Test: Searching for Breakpoint — Total Electricity")
lines(breakpoints(qlr_total))
```



## Saving the data


```{r save-data}
# Save as CSV files in your working directory
write_csv(
  total_elec %>% as_tibble(),
  "data/total_electricity_norway.csv"
)

write_csv(
  hydro %>% as_tibble(),
  "data/hydro_norway.csv"
)

cat("Files saved successfully!\n")
cat("- total_electricity_norway.csv\n")
cat("- hydro_norway.csv\n")
```
