---
title: "Predictive Analytics"
subtitle: "Final Project: Norwegian Hydroelectric Production x Reservoir Fill Rate"

author:
  - "Carla de Erausquin (185899)"
  - "Christoph Schilling (185939)"
  - "Florencia Brazionis (175974)"
  - "José Miguel Gomes Queirós Lobo (185920)"

date: "`r format(Sys.Date(), '%d %B %Y')`"

output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    toc: true
    toc_depth: 2
    number_sections: false

geometry:
  - a4paper
  - top=20mm
  - left=15mm
  - right=15mm
  - bottom=20mm

header-includes:
  - \usepackage{titlesec}
  - \usepackage{titling}
  - \usepackage{xcolor}
  - \pretitle{\begin{center}\vspace*{\fill}\LARGE\bfseries}
  - \posttitle{\end{center}\vspace*{\fill}}
  - \renewcommand{\maketitlehookd}{\thispagestyle{empty}\clearpage}
  - \let\oldtoc\tableofcontents
  - \renewcommand{\tableofcontents}{\oldtoc\clearpage}
---

# Setup

First, we load the necessary packages and set the working directory.

```{r setup, message=FALSE, warning=FALSE}
# Clear the workspace
rm(list = ls())

# 1. Install pak if not already available (pak itself is on CRAN)
if (!requireNamespace("pak", quietly = TRUE)) install.packages("pak")

# ── Data import & wrangling ───────────────────────────────────────────────────
# readr:    reads CSV files with read_csv()
# jsonlite: parses JSON responses from the NVE and SSB APIs via fromJSON()
# stringr:  string manipulation via str_replace() and related functions
# ── Time series & forecasting ─────────────────────────────────────────────────
# fpp3:     meta-package for the course — loads tsibble, fable, feasts,
#           ggplot2, dplyr, and lubridate in one call
# seasonal: interface to X-13ARIMA-SEATS for seasonal decomposition
# tseries:  unit root testing via adf.test()
# urca:     unit root and cointegration tests via ur.kpss() and ur.df()
# ── Visualisation ─────────────────────────────────────────────────────────────
# waffle:   waffle chart geometry for the electricity generation mix plot
#           (not on CRAN — installed from GitHub)
# ggtext:   renders HTML in ggplot2 text elements for colour-coded annotations
# patchwork: combines multiple ggplot2 plots into a single figure
# ── Report generation ─────────────────────────────────────────────────────────
# knitr:    renders the R Markdown notebook to PDF or HTML

# 2. Define all packages
# pak accepts CRAN names and "user/repo" GitHub format in the same vector
packages <- c(
  # Data import & wrangling
  "readr",
  "jsonlite",
  "stringr",
  # Time series & forecasting
  "fpp3",
  "seasonal",
  "tseries",
  "urca",
  # Visualisation
  "hrbrmstr/waffle",  # GitHub — not on CRAN
  "ggtext",
  "patchwork",
  # Report generation
  "knitr"
)

# 3. Install all missing packages
# pak skips already-installed packages automatically
pak::pak(packages)

# 4. Load libraries
# Strip "user/" GitHub prefix — library() only accepts the package name
library_names <- sub(".*/", "", packages)
invisible(lapply(library_names, library, character.only = TRUE))

# 5. Confirm all packages loaded successfully
missing <- library_names[!library_names %in% loadedNamespaces()]
if (length(missing) > 0) {
  message("The following packages failed to load: ", paste(missing, collapse = ", "))
} else {
  message("All packages installed and loaded successfully.")
}
```


# Data Loading

Then, we format the data into a `tsibble` (Time Series Tibble), ensuring the 
date is parsed correctly.

```{r load_data_inflow}
# ==============================================================================
# File 1: Reservoir Inflow — Weekly, 1958–2026, GWh
# Source: NVE (2026), https://www.nve.no/energi/analyser-og-statistikk/hydrologiske-data-til-kraftsituasjonsrapporten/
# ==============================================================================
# NB: Useful inflow is drawn from NVE's production-based calculation (derived 
# from turbine metering and reservoir level measurements) from 2000 onwards, and 
# from the HBV hydrological model for earlier observations. We combine them into 
# one unified series. The production-based method (adopted in 2015 and back-
# -calculated until 2000) is preferred as it relies on direct measurements 
# rather than simulated estimates; the HBV series is used solely to extend the 
# descriptive history prior to 2000 and is not used in model estimation.
# ──────────────────────────────────────────────────────────────────────────────

# 1. Load data
raw_inflow <- read_csv("data/NVE_ReservoirInflow_1958-2026_GWh.csv",
                       show_col_types = FALSE)

# Inspect
glimpse(raw_inflow)

# 2. Filter and transmute relevant data and convert to a tsibble
inflow <- raw_inflow %>%
  filter(`Område` == "NO") %>%    # national aggregate only
  rename(
    inflow_post2000_GWh = `Nyttbart tilsig (produksjonsdata og magasinstatistikk, brukes f.o.m. 2015)`,
    inflow_pre2000_GWh  = `Nyttbart tilsig HBV (brukes før 2015 og til min/maks/gj.snitt)`
  ) %>%
  transmute(
    # Build a standard date from year + week number
    year_week = yearweek(paste0(`År`, " W", sprintf("%02d", `Uke`))),
    # Combine the two inflow methods into one column (primary regressor):
    # production-based is preferred when available (2000 onwards),
    # HBV model is used for all earlier observations and recent weeks
    # where production-based has not yet been published.
    useful_inflow_GWh = if_else(
      !is.na(inflow_post2000_GWh),
      inflow_post2000_GWh,    # production-based where available (2000–present)
      inflow_pre2000_GWh      # HBV fallback for 1958–1999
    ),
    precip_energy_GWh      = `Nedbørsenergi`, # useful for data section description
    runoff_groundwater_GWh = `Mark-og grunnvann`, # idem
    snow_GWh               = `Snø` # snowpack — useful for seasonal narrative
  ) %>%
  arrange(year_week) %>%      # sort ascending (raw file is descending)
  as_tsibble(index = year_week)

# Inspect
glimpse(inflow)
```

```{r load_data_fill_rate}
# ==============================================================================
# File 2: Reservoir Fill Rate – Weekly, 1994-2026, TWh (-> GWh)
# Source: NVE (2026), https://api.nve.no/doc/magasinstatistikk/
# ==============================================================================

# 1. Download API response
url_main <- paste0(
  "https://biapi.nve.no/magasinstatistikk/api/Magasinstatistikk",
  "/HentOffentligData"
)
url_ref  <- paste0(
  "https://biapi.nve.no/magasinstatistikk/api/Magasinstatistikk",
  "/HentOffentligDataMinMaxMedian"
)

raw_fill_rate <- fromJSON(url_main)
raw_fill_ref  <- fromJSON(url_ref)

# Inspect
glimpse(raw_fill_rate)

# 2. Filter and transmute relevant data and convert to a tsibble
fill_rate <- raw_fill_rate %>%
  filter(omrType == "NO") %>%                # NO = Norway (whole country)
  transmute(
    year_week        = yearweek(as.Date(dato_Id)),
    fill_rate        = fyllingsgrad,
    energy_GWh       = fylling_TWh * 1000,   # stored energy content
    capacity_GWh     = kapasitet_TWh * 1000, # total system capacity
    change_prev_week = endring_fyllingsgrad  # weekly delta (kept for reference)
  ) %>%
  arrange(year_week) %>%      # sort ascending (raw file is unordered)
  as_tsibble(index = year_week)

# Inspect
glimpse(fill_rate)
```

```{r load_data_electricity}
# ══════════════════════════════════════════════════════════════════════════════
# File 3: Electricity balance, by production and consumption and month, 
#         1993–2026, MWh (-> GWh)
# Source: SSB (2026), table 14091, https://www.ssb.no/en/statbank/table/14091
# ══════════════════════════════════════════════════════════════════════════════

# Each '&' introduces a query parameter telling the SSB API what to return
ssb_url <- paste0(
  "https://data.ssb.no/api/pxwebapi/v2/tables/14091/data",
  "?lang=en",
  "&outputFormat=json-stat2",
  "&valuecodes[ContentsCode]=*",
  "&valuecodes[Tid]=*",
  "&valuecodes[Produk2]=1,1.1,1.2,1.3,1.4,2,3,4",
  "&heading=ContentsCode,Tid",
  "&stub=Produk2"
)

# 1. Download API response and parse the JSON into an R list. 
# simplifyVector = TRUE automatically converts simple JSON arrays into R vectors 
# rather than lists, making the data easier to work with.
ssb_json <- fromJSON(ssb_url, simplifyVector = TRUE)

# 2. Extract dimension metadata
# Produk2: 8 categories
#   1. Total electricity generation
#     1.1. Hydro power generation
#     1.2. Wind power generation
#     1.3. Solar power generation
#     1.4. Thermal power generation
#   2. Import
#   3. Export
#   4. Gross electricity consumption
# Tid: 399 months (1993M01 to 2026M03)
n_categories <- ssb_json$size[1]    # 8
n_months     <- ssb_json$size[3]    # 399

cat_codes  <- names(ssb_json$dimension$Produk2$category$index)  # "1","1.1",...
cat_labels <- unlist(ssb_json$dimension$Produk2$category$label) # readable names
month_strs <- names(ssb_json$dimension$Tid$category$index)      # "1993M01",...

# 3. Reconstruct long-format dataframe
# JSON-stat2 value array layout: Produk2 varies slowest, Tid varies fastest.
# Each of the 8 categories occupies a contiguous block of 399 monthly values.
raw_electricity <- tibble(
  category_code  = rep(cat_codes,  each  = n_months),
  category_label = rep(cat_labels, each  = n_months),
  month_str      = rep(month_strs, times = n_categories),
  value_MWh      = ssb_json$value    # NAs where SSB reports ".."
)

# Inspect
glimpse(raw_electricity)

# 4. Create wide view for inspection only
electricity_wide <- raw_electricity %>%
  pivot_wider(
    names_from  = category_label,
    values_from = value_MWh,
    id_cols     = month_str
  )

# Inspect
glimpse(electricity_wide)

# 5. Create tidy tsibble for analysis
electricity <- raw_electricity %>%
  mutate(
    value_GWh  = value_MWh / 1000,                        # convert MWh to GWh
    date       = ym(str_replace(month_str, "M", "-")),    # "1993M01" -> date
    year_month = yearmonth(date)
  ) %>%
  select(year_month, category_code, category_label, value_GWh) %>%
  as_tsibble(index = year_month, key = category_code) %>%
  arrange(year_month, category_code)

# Inspect
glimpse(electricity)

# 6. Create hydro-specific tsibble for modelling
hydro <- electricity %>%
  filter(category_code == "1.1") %>%
  select(year_month, hydro_GWh = value_GWh)

# Inspect
glimpse(hydro)
```


# Exploratory Data Analysis

```{r EDA}
# ══════════════════════════════════════════════════════════════════════════════
# EXPLORATORY ANALYSIS: hydro_GWh
# ══════════════════════════════════════════════════════════════════════════════

# ── 1. autoplot() — full time series ─────────────────────────────────────────
# What to look for: trend, variance changes, outliers, structural breaks
hydro %>%
  autoplot(hydro_GWh) +
  labs(
    title   = "Norwegian hydroelectric production — monthly (1993–2026)",
    y       = "Production (GWh)",
    x       = NULL,
    caption = "Source: SSB, table 14091"
  )

# ── 2. gg_season() — seasonal plot ───────────────────────────────────────────
# What to look for: consistent seasonal pattern across years, anomalous years
# Each line = one year. Expect: peak in winter (Jan/Feb), trough in summer
hydro %>%
  gg_season(hydro_GWh, labels = "right") +
  labs(
    title    = "Seasonal plot: Norwegian hydroelectric production",
    subtitle = "Each line represents one calendar year",
    y        = "Production (GWh)",
    x        = NULL
  )

# ── 3. gg_subseries() — subseries plot ───────────────────────────────────────
# What to look for: within-month trend over years, whether seasonality is stable
# Each panel = one month. Blue horizontal line = mean for that month.
# Expect: January and February panels to sit higher than June/July
hydro %>%
  gg_subseries(hydro_GWh) +
  labs(
    title    = "Subseries plot: Norwegian hydroelectric production",
    subtitle = "Each panel shows one month across all years. Blue line = monthly mean.",
    y        = "Production (GWh)",
    x        = NULL
  )

# ── 4. gg_lag() — lag plots ───────────────────────────────────────────────────
# What to look for: linear structure at lag 12 confirms annual seasonality;
# linear structure at lag 1 confirms autocorrelation; scatter = no relationship
hydro %>%
  gg_lag(hydro_GWh, lags = 1:12, geom = "point") +
  labs(
    title    = "Lag plots: Norwegian hydroelectric production",
    subtitle = "Strong linear pattern at lag 12 confirms annual seasonality"
  )

# ── 5. ACF() — autocorrelation function ───────────────────────────────────────
# What to look for:
#   - Slow decay → non-stationary (needs differencing)
#   - Spikes at lag 12, 24 → seasonal autocorrelation
#   - Sharp cut-off → MA structure
#   - Gradual decay → AR structure
hydro %>%
  ACF(hydro_GWh, lag_max = 48) %>%   # 48 lags = 4 years
  autoplot() +
  labs(
    title    = "ACF: Norwegian hydroelectric production",
    subtitle = "Dashed lines = 95% significance bounds",
    x        = "Lag (months)"
  )

# ── 5b. ACF after first difference — check if differencing induces stationarity
hydro %>%
  mutate(d_hydro_GWh = difference(hydro_GWh)) %>%
  ACF(d_hydro_GWh, lag_max = 48) %>%
  autoplot() +
  labs(
    title    = "ACF: first-differenced hydroelectric production",
    subtitle = "Residual spikes at lag 12, 24 indicate remaining seasonal structure",
    x        = "Lag (months)"
  )

# ── 5c. ACF after seasonal difference — check if seasonal differencing helps
hydro %>%
  mutate(ds_hydro_GWh = difference(hydro_GWh, lag = 12)) %>%
  ACF(ds_hydro_GWh, lag_max = 48) %>%
  autoplot() +
  labs(
    title    = "ACF: seasonally differenced hydroelectric production",
    subtitle = "Lag = 12 months",
    x        = "Lag (months)"
  )
```


# Reservoir Data — Monthly Aggregation and Join
```{r reservoir_monthly}
# ==============================================================================
# Join weekly reservoir data (inflow + fill rate) with monthly hydro production
# Strategy: aggregate weekly → monthly, then left-join on year_month
# ==============================================================================

# 1. Aggregate inflow to monthly (sum weekly inflows within each month)
# Inflow is a flow variable — it measures a quantity accumulating over time.
# Therefore, we sum the values.

inflow_monthly <- inflow %>%
  index_by(year_month = yearmonth(year_week)) %>%
  summarise(
    inflow_GWh             = sum(useful_inflow_GWh,      na.rm = TRUE),
    precip_energy_GWh      = sum(precip_energy_GWh,      na.rm = TRUE),
    runoff_groundwater_GWh = sum(runoff_groundwater_GWh, na.rm = TRUE),
    snow_GWh               = sum(snow_GWh,               na.rm = TRUE)
  )

# 2. Aggregate fill rate to monthly (mean level within each month)
# Fill rate is a stock variable — it measures a level at a point in time. The monthly
# fill rate should represent the average state of the reservoir system over the month, 
# not the total. 
# Therefore, we compute the mean – the standard choice for stock variables aggregated
# to lower frequency.

fill_rate_monthly <- fill_rate %>%
  index_by(year_month = yearmonth(year_week)) %>%
  summarise(
    fill_rate_mean  = mean(fill_rate,  na.rm = TRUE),
    energy_GWh_mean = mean(energy_GWh, na.rm = TRUE),
    capacity_GWh    = first(capacity_GWh) # constant: identical across all weeks
  )

# 3. Join with hydro production on year_month
reservoir <- hydro %>%
  inner_join(inflow_monthly,    by = "year_month") %>%
  inner_join(fill_rate_monthly, by = "year_month") %>%
  as_tsibble(index = year_month)

# Inspect
glimpse(reservoir)
```


# EDA 2
```{r}
# ══════════════════════════════════════════════════════════════════════════════
# WAFFLE CHART: Norway's electricity generation mix — annual, 1993–2025
# Adapted from R Graph Gallery tutorial (Guillaume Noblet)
# Source data: electricity tsibble (SSB table 14091)
# ══════════════════════════════════════════════════════════════════════════════

# ── 1. Prepare data ───────────────────────────────────────────────────────────

# Define sources and colours
# Norway's mix is almost entirely hydro — colours reflect Norwegian energy identity
energy_sources <- c(
  "Hydro"   = "1.1",
  "Wind"    = "1.2",
  "Solar"   = "1.3",
  "Thermal" = "1.4"
)

energy_colours <- c(
  "Hydro"   = "#3A7DC9",   # deep blue — water
  "Wind"    = "#A8D5E2",   # light blue — air
  "Solar"   = "#F8D564",   # yellow — sun
  "Thermal" = "#C0856A"    # warm brown — combustion
)

# Aggregate to annual GWh per source
# NA treated as 0 (wind and solar not reported in early years = truly zero)
waffle_data <- electricity %>%
  as_tibble() %>%
  filter(
    category_code %in% energy_sources,
    lubridate::year(as.Date(year_month)) >= 2009   # start from 2009
  ) %>%
  mutate(
    year   = lubridate::year(as.Date(year_month)),
    source = names(energy_sources)[match(category_code, energy_sources)],
    source = factor(source, levels = names(energy_sources))
  ) %>%
  group_by(year, source) %>%
  summarise(
    annual_GWh = sum(value_GWh, na.rm = TRUE),
    .groups    = "drop"
  )

# Compute percentage of each source within each year
waffle_data <- waffle_data %>%
  group_by(year) %>%
  mutate(perc = (annual_GWh / sum(annual_GWh)) * 100) %>%
  ungroup()

# Round to integers — simple rounding may not sum to exactly 100
# Apply correction to the largest category (same logic as tutorial)
waffle_data <- waffle_data %>%
  group_by(year) %>%
  mutate(
    perc_int = as.integer(round(perc)),
    perc_int = {
      diff <- 100L - sum(perc_int)
      if (diff != 0L) perc_int[which.max(annual_GWh)] <-
        perc_int[which.max(annual_GWh)] + diff
      perc_int
    }
  ) %>%
  filter(perc_int > 0) %>%    # drop source-years with 0 squares (Solar pre-2010)
  ungroup()

# ── 2. Build subtitle with colour-coded source labels ────────────────────────
subtitle_text <- glue::glue(
  "<span style='color:{energy_colours['Hydro']};'><strong>Hydro</strong></span>
  has dominated Norwegian electricity generation since records began.
  <span style='color:{energy_colours['Wind']};'><strong>Wind</strong></span>
  began entering the mix in the 2000s and has grown steadily,
  while <span style='color:{energy_colours['Thermal']};'><strong>thermal</strong></span>
  and <span style='color:{energy_colours['Solar']};'><strong>solar</strong></span>
  remain marginal. Each square represents 1% of annual generation."
)

# ── 3. Plot ───────────────────────────────────────────────────────────────────
# n_rows = 10 → each waffle is 10 × 10 squares (compact, manageable for 33 years)
# flip = TRUE → squares fill vertically (bottom to top)
g <- ggplot(waffle_data, aes(fill = source, values = perc_int)) +
  geom_waffle(
    colour           = "white",
    size             = 0.3,
    n_rows           = 10,
    flip             = TRUE,
    make_proportional = FALSE
  ) +
  facet_wrap(
    ~year,
    nrow           = 1,
    strip.position = "bottom"
  ) +
  scale_x_discrete() +
  scale_y_continuous(
    labels = function(x) x * 10,    # each row = 10% (100 squares / 10 rows)
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = energy_colours,
    breaks = names(energy_sources),
    name   = "Source",
    drop   = FALSE
  ) +
  theme_minimal(base_family = body_font) +
  labs(
    title    = "Norway's Electricity Generation Mix (1993–2025)",
    subtitle = subtitle_text,
    caption  = "Each square = 1% of total annual generation. Sources: SSB table 14091."
  ) +
  theme(
    panel.grid      = element_blank(),
    axis.title      = element_blank(),
    axis.text.x     = element_blank(),   # year shown in strip label below
    axis.text.y     = element_blank(),
    strip.text      = element_text(
      family = body_font,
      size   = 7,
      angle  = 90,                       # rotate year labels — 33 years is tight
      vjust  = 0.5
    ),
    legend.position = "bottom",
    legend.title    = element_text(family = body_font, size = 10),
    legend.text     = element_text(family = body_font, size = 10),
    plot.title      = element_textbox_simple(
      hjust  = 0,
      margin = margin(20, 0, 10, 0),
      size   = 18,
      family = title_font,
      face   = "bold",
      color  = "grey15",
      width  = unit(0.9, "npc")
    ),
    plot.subtitle   = element_textbox_simple(
      hjust  = 0,
      margin = margin(10, 0, 30, 0),
      width  = unit(0.9, "npc"),
      size   = 12,
      family = body_font,
      color  = "grey15"
    ),
    plot.caption    = element_textbox_simple(
      family = body_font,
      size   = 9,
      color  = "grey40",
      hjust  = 0,
      width  = unit(0.95, "npc"),
      margin = margin(10, 0, 0, 0)
    ),
    plot.background = element_rect(color = "white", fill = "white"),
    plot.margin     = margin(20, 20, 20, 20)
  )

g

# ── 4. Save — wide aspect ratio essential for 33 annual columns ───────────────
ggsave(
  "waffle_norway_electricity_mix.png",
  plot   = g,
  width  = 24,     # wide to accommodate 33 years
  height = 6,
  dpi    = 300,
  bg     = "white"
)
```


```{r eda 2}
# ══════════════════════════════════════════════════════════════════════════════
# STACKED AREA CHART: Norway's electricity generation by source — annual
# Source data: electricity tsibble (SSB table 14091)
# ══════════════════════════════════════════════════════════════════════════════

# ── 1. Prepare data ───────────────────────────────────────────────────────────
area_data <- electricity %>%
  as_tibble() %>%
  filter(category_code %in% energy_sources) %>%
  mutate(
    year   = lubridate::year(as.Date(year_month)),
    source = names(energy_sources)[match(category_code, energy_sources)],
    source = factor(source, levels = rev(names(energy_sources)))
  ) %>%
  group_by(year, source) %>%
  summarise(
    annual_GWh = sum(value_GWh, na.rm = TRUE),
    .groups    = "drop"
  )

# ── 2. Plot A: absolute GWh — shows total volume AND composition ──────────────
p_absolute <- ggplot(area_data, aes(x = year, y = annual_GWh, fill = source)) +
  geom_area(colour = "white", linewidth = 0.2, alpha = 0.9) +
  scale_fill_manual(
    values = energy_colours,
    breaks = names(energy_sources),
    name   = "Source"
  ) +
  scale_x_continuous(breaks = seq(1993, 2025, by = 4), expand = c(0, 0)) +
  scale_y_continuous(
    labels = scales::comma_format(suffix = " GWh"),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  labs(
    title    = "Norwegian Electricity Generation by Source (1993–2025)",
    subtitle = "Annual production in GWh. Wind and solar are negligible before 2005.",
    y        = "Annual production (GWh)",
    x        = NULL,
    caption  = "Source: SSB table 14091."
  ) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position    = "bottom",
    plot.background    = element_rect(color = "white", fill = "white"),
    plot.margin        = margin(20, 20, 20, 20)
  )

p_absolute

# ── 3. Plot B: 100% stacked — shows composition only ─────────────────────────
p_percent <- ggplot(area_data, aes(x = year, y = annual_GWh, fill = source)) +
  geom_area(
    colour    = "white",
    linewidth = 0.2,
    alpha     = 0.9,
    position  = "fill"
  ) +
  scale_fill_manual(
    values = energy_colours,
    breaks = names(energy_sources),
    name   = "Source"
  ) +
  scale_x_continuous(breaks = seq(1993, 2025, by = 4), expand = c(0, 0)) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = c(0, 0)
  ) +
  theme_minimal() +
  labs(
    title    = "Share of Norwegian Electricity Generation by Source (1993–2025)",
    subtitle = "Proportional composition. Hydro dominates; wind share has grown since 2005.",
    y        = "Share of annual generation (%)",
    x        = NULL,
    caption  = "Source: SSB table 14091."
  ) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position    = "bottom",
    plot.background    = element_rect(color = "white", fill = "white"),
    plot.margin        = margin(20, 20, 20, 20)
  )

p_percent

# ── 4. Save ───────────────────────────────────────────────────────────────────
ggsave("stacked_area_absolute.png", plot = p_absolute,
       width = 10, height = 6, dpi = 300, bg = "white")

ggsave("stacked_area_percent.png",  plot = p_percent,
       width = 10, height = 6, dpi = 300, bg = "white")
```



# Sanity Checks

```{r sanity_checks}

# ── File 1: inflow ────────────────────────────────────────────────────────────

# 1a. Row count — expect ~3,536 rows (68 years × 52 weeks, 1958–2026)
cat("inflow rows:", nrow(inflow), "\n")

# 1b. Date range — expect 1958 W01 to 2026 W(latest)
cat("inflow range:", as.character(min(inflow$year_week)),
    "→", as.character(max(inflow$year_week)), "\n")

# 1c. Gaps in weekly index
cat("inflow has gaps:", has_gaps(inflow)[[1]], "\n")
count_gaps(inflow)

# 1d. Completely empty rows (all key measurement columns NA simultaneously)
inflow %>%
  filter(
    is.na(useful_inflow_GWh),
    is.na(precip_energy_GWh),
    is.na(snow_GWh),
    is.na(runoff_groundwater_GWh)
  ) %>%
  nrow() %>%
  { cat("inflow completely empty rows:", ., "\n") }

# 1e. NA count per column
cat("\n--- inflow: NA counts per column ---\n")
inflow %>%
  as_tibble() %>%
  summarise(across(everything(), \(x) sum(is.na(x)))) %>%
  glimpse()

# 1f. Verify if_else logic: how many rows use each inflow source
# Expect: production-based (post2000) covers ~2000-2026, HBV covers 1958-1999
cat("\n--- inflow: source breakdown ---\n")
raw_inflow %>%
  filter(`Område` == "NO") %>%
  rename(
    prod = `Nyttbart tilsig (produksjonsdata og magasinstatistikk, brukes f.o.m. 2015)`,
    hbv  = `Nyttbart tilsig HBV (brukes før 2015 og til min/maks/gj.snitt)`
  ) %>%
  summarise(
    n_prod_based    = sum(!is.na(prod)),   # rows where post2000 is available
    n_hbv_fallback  = sum(is.na(prod)),    # rows where HBV is used
    n_both_na       = sum(is.na(prod) & is.na(hbv))  # should be 0
  ) %>%
  print()

# 1g. Physical plausibility: inflow must be non-negative
n_neg <- inflow %>% filter(useful_inflow_GWh < 0) %>% nrow()
cat("inflow negative values:", n_neg, "(expect 0)\n")

# 1h. Summary statistics
cat("\n--- inflow: summary of useful_inflow_GWh ---\n")
summary(inflow$useful_inflow_GWh)

# 1i. Visual check — strong seasonality expected
inflow %>%
  autoplot(useful_inflow_GWh) +
  labs(title = "Norwegian reservoir inflow — weekly (1958–2026)",
       y = "Inflow (GWh)", x = NULL)


# ── File 2: fill_rate ─────────────────────────────────────────────────────────

# 2a. Critical check first: did the filter return any rows?
# If fill_rate has 0 rows, omrType == "NO" is wrong — correct value is "L"
cat("fill_rate rows:", nrow(fill_rate), "(expect ~1,664; if 0, check omrType filter)\n")

# 2b. What omrType values actually exist in the raw data?
cat("\n--- fill_rate: distinct omrType values in raw data ---\n")
raw_fill_rate %>%
  distinct(omrType, omrnr) %>%
  arrange(omrType, omrnr) %>%
  print()

# 2c. Date range — expect 1994/1995 W(earliest) → 2026 W(latest)
cat("fill_rate range:", as.character(min(fill_rate$year_week)),
    "→", as.character(max(fill_rate$year_week)), "\n")

# 2d. Gaps in weekly index
cat("fill_rate has gaps:", has_gaps(fill_rate)[[1]], "\n")
count_gaps(fill_rate)

# 2e. Completely empty rows
fill_rate %>%
  filter(is.na(fill_rate), is.na(energy_GWh), is.na(capacity_GWh)) %>%
  nrow() %>%
  { cat("fill_rate completely empty rows:", ., "\n") }

# 2f. NA count per column
cat("\n--- fill_rate: NA counts per column ---\n")
fill_rate %>%
  as_tibble() %>%
  summarise(across(everything(), \(x) sum(is.na(x)))) %>%
  glimpse()

# 2g. Physical range: fill_rate should be between 0 and 1
# (stored as proportion, not percentage)
cat("\n--- fill_rate: value range ---\n")
fill_rate %>%
  summarise(
    min_fill     = min(fill_rate,    na.rm = TRUE),
    max_fill     = max(fill_rate,    na.rm = TRUE),
    min_energy   = min(energy_GWh,   na.rm = TRUE),
    max_energy   = max(energy_GWh,   na.rm = TRUE),
    min_capacity = min(capacity_GWh, na.rm = TRUE),
    max_capacity = max(capacity_GWh, na.rm = TRUE)
  ) %>%
  print()

# 2h. capacity_GWh should be positive and slowly increasing over time
# (new plants come online — step increases are expected)
fill_rate %>%
  as_tibble() %>%
  mutate(year = lubridate::year(as.Date(year_week))) %>%
  group_by(year) %>%
  summarise(mean_capacity_GWh = mean(capacity_GWh, na.rm = TRUE)) %>%
  print(n = Inf)

# 2i. Visual checks
fill_rate %>%
  autoplot(fill_rate) +
  labs(title = "Norwegian reservoir fill rate — weekly",
       subtitle = "Expect: trough in spring, peak in autumn",
       y = "Fill rate (proportion)", x = NULL)

fill_rate %>%
  autoplot(energy_GWh) +
  labs(title = "Norwegian reservoir energy content — weekly",
       y = "Energy (GWh)", x = NULL)


# ── File 3: electricity ───────────────────────────────────────────────────────

# 3a. Row count — expect 8 categories × 399 months = 3,192 rows
cat("electricity rows:", nrow(electricity),
    "(expect 3,192 =", n_categories, "×", n_months, ")\n")

# 3b. All 8 category codes present
cat("\n--- electricity: categories present ---\n")
electricity %>%
  as_tibble() %>%
  distinct(category_code, category_label) %>%
  arrange(category_code) %>%
  print()

# 3c. Date range — expect 1993 Jan → 2026 Mar
cat("electricity range:", as.character(min(electricity$year_month)),
    "→", as.character(max(electricity$year_month)), "\n")

# 3d. Gaps per category (each should have 399 months, no gaps)
cat("\n--- electricity: gaps per category ---\n")
electricity %>%
  group_by(category_code) %>%
  summarise(
    n_rows  = n(),
    n_gaps  = sum(is.na(value_GWh)),
    min_ym  = as.character(min(year_month)),
    max_ym  = as.character(max(year_month))
  ) %>%
  print()

# 3e. NA pattern — wind (1.2) and solar (1.3) expected to have many early NAs
cat("\n--- electricity: NA counts per category ---\n")
electricity %>%
  as_tibble() %>%
  group_by(category_code, category_label) %>%
  summarise(
    n_total   = n(),
    n_na      = sum(is.na(value_GWh)),
    pct_na    = round(100 * n_na / n_total, 1),
    first_non_na = as.character(
      min(year_month[!is.na(value_GWh)], na.rm = TRUE)
    ),
    .groups = "drop"
  ) %>%
  print()

# 3f. Physical plausibility: no negative generation values
cat("\n--- electricity: negative values per category ---\n")
electricity %>%
  as_tibble() %>%
  filter(!is.na(value_GWh), value_GWh < 0) %>%
  group_by(category_code, category_label) %>%
  summarise(n_negative = n(), .groups = "drop") %>%
  { if (nrow(.) == 0) cat("No negative values found (expected)\n") else print(.) }

# 3g. Internal consistency check:
# Total (1) should equal sum of 1.1 + 1.2 + 1.3 + 1.4 for each month
# Tolerance of 1 GWh for rounding
cat("\n--- electricity: total vs sum of sub-categories (first 10 mismatches) ---\n")
electricity %>%
  as_tibble() %>%
  filter(category_code %in% c("1", "1.1", "1.2", "1.3", "1.4")) %>%
  group_by(year_month) %>%
  summarise(
    total    = sum(value_GWh[category_code == "1"],   na.rm = TRUE),
    sum_sub  = sum(value_GWh[category_code != "1"],   na.rm = TRUE),
    diff_GWh = abs(total - sum_sub),
    .groups  = "drop"
  ) %>%
  filter(diff_GWh > 1) %>%          # flag months where they disagree by > 1 GWh
  arrange(desc(diff_GWh)) %>%
  head(10) %>%
  print()

# 3h. Summary statistics for hydro (primary modelling series)
cat("\n--- hydro: summary ---\n")
summary(hydro$hydro_GWh)

# 3i. Visual checks
hydro %>%
  autoplot(hydro_GWh) +
  labs(title = "Norwegian hydroelectric production — monthly (1993–2026)",
       y = "Production (GWh)", x = NULL)

electricity %>%
  filter(category_code %in% c("1.1", "1.2", "1.3")) %>%
  autoplot(value_GWh) +
  facet_wrap(~category_label, ncol = 1, scales = "free_y") +
  labs(title = "Norwegian electricity generation by source — monthly",
       y = "GWh", x = NULL)
```

NB: Norwegian hydro capacity has grown over time, so we should consider using 
production per unit capacity (or log-transforming) to control for the 
infrastructure expansion trend.


OLD

# # ── Sanity checks ─────────────────────────────────────────────────────────────
#
# # Expect 399 rows (1993M01 → 2026M03)
# nrow(ssb_hydro)
# range(ssb_hydro$date)
# has_gaps(ssb_hydro)
#
# ssb_hydro %>%
#   summarise(na_hydro = sum(is.na(hydro_GWh)))
#
# ssb_hydro %>%
#   autoplot(hydro_GWh) +
#   labs(title = "Norwegian hydroelectric production — monthly (1993–2026)",
#        y = "Production (GWh)", x = NULL)
#
# # ══════════════════════════════════════════════════════════════════════════════
# # QUICK SANITY CHECKS
# # ══════════════════════════════════════════════════════════════════════════════
#
# # NVE: expect ~3,500 weekly rows (1958–2026 ≈ 68 years × 52 weeks)
# nrow(inflow)
# range(inflow$year)
# has_gaps(inflow)
#
# # SSB: expect ~398 monthly rows (1993M01–2026M03 = 33 years × 12 months + 3)
# nrow(ssb_hydro)
# range(ssb_hydro$date)
# has_gaps(ssb_hydro)
#
# # Check for NAs in the key columns
# inflow %>%
#   summarise(
#     na_useful_inflow  = sum(is.na(useful_inflow_GWh)),
#     na_precip         = sum(is.na(precip_energy_GWh))
#   )
#
# ssb_hydro %>%
#   summarise(na_hydro = sum(is.na(hydro_GWh)))
#
# # Visual check — seasonal patterns should be immediately visible
# inflow %>%
#   autoplot(useful_inflow_GWh) +
#   labs(title = "Norwegian useful reservoir inflow — weekly (1958–2026)",
#        y = "Inflow (GWh)", x = NULL)
#
# ssb_hydro %>%
#   autoplot(hydro_GWh) +
#   labs(title = "Norwegian hydroelectric production — monthly (1993–2026)",
#        y = "Production (GWh)", x = NULL)
#
#
#
#
# # ── 5. SANITY CHECKS (File 2) ─────────────────────────────────────────────────
#
# # 5a. How many weeks total? Expect ~30 years × 52 weeks ≈ 1,560 rows
# nrow(fill_rate)
#
# # 5b. What years are covered?
# range(fill_rate$year_week)
#
# # 5c. Any gaps in the weekly sequence? (has_gaps detects missing index rows)
# has_gaps(fill_rate)
# count_gaps(fill_rate)
#
# # 5d. Any NA values in key columns?
# fill_rate %>%
#   summarise(
#     na_fill     = sum(is.na(fill_rate)),
#     na_energy   = sum(is.na(energy_GWh)),
#     na_capacity = sum(is.na(capacity_GWh))
#   )
#
# # 5e. Are fill rate values within the valid 0–100% range?
# fill_rate %>%
#   summarise(
#     min_fill = min(fill_rate, na.rm = TRUE),
#     max_fill = max(fill_rate, na.rm = TRUE)
#   )
#
# # 5f. Quick visual check — does the seasonal pattern look right?
# # Expect: peak in autumn (~Q4), trough in spring (~Q1/Q2)
# fill_rate %>%
#   autoplot(fill_rate) +
#   labs(title = "Norwegian reservoir fill rate — weekly (1995–present)",
#        y = "Fill rate", x = NULL)
#
# # 5g. Energy content over time (should broadly track fill rate)
# fill_rate %>%
#   autoplot(energy_GWh) +
#   labs(title = "Norwegian reservoir energy content — weekly",
#        y = "Energy (GWh)", x = NULL)

# ── 6. DESCRIPTIVE PLOT: historical envelope with selected years ───────────────
# Uses HentOffentligDataMinMaxMedian to build a min/max/median ribbon by week
# number, then overlays selected individual years on top. Choose years that
# tell a story relevant to your analysis:
#   - One anomalously dry year (2022 tracked near historical minimum throughout,
#     driving record spot prices — directly motivates fill rate as a regressor)
#   - One near-median year (2020 illustrates the normal seasonal cycle)
#   - One recent year (2024 shows whether the system has recovered)
# This figure belongs in the Data section. It simultaneously demonstrates
# seasonality in fill rate, its bounded/stationary nature, and the inter-annual
# variation that gives your regressor explanatory power.

# envelope <- raw_fill_ref %>%
#   filter(omrType == "L") %>%
#   transmute(
#     week        = uke,
#     min_fill    = minFyllingsgrad * 100,
#     max_fill    = maxFyllingsgrad * 100,
#     median_fill = medianFyllingsgrad * 100
#   )
#
# selected_years <- raw_fill_rate %>%
#   filter(omrType == "L", aar %in% c(2020, 2022, 2024)) %>%
#   transmute(
#     week     = uke,
#     year     = factor(aar),
#     fill_pct = fyllingsgrad * 100
#   )
#
# ggplot() +
#   geom_ribbon(
#     data = envelope,
#     aes(x = week, ymin = min_fill, ymax = max_fill),
#     fill = "steelblue", alpha = 0.15
#   ) +
#   geom_line(
#     data = envelope,
#     aes(x = week, y = median_fill),
#     colour = "steelblue", linewidth = 0.6, linetype = "dashed"
#   ) +
#   geom_line(
#     data = selected_years,
#     aes(x = week, y = fill_pct, colour = year),
#     linewidth = 0.9
#   ) +
#   scale_colour_manual(
#     values = c("2020" = "#2ecc71", "2022" = "#e74c3c", "2024" = "#f39c12"),
#     name   = "Year"
#   ) +
#   scale_x_continuous(
#     breaks = c(1, 13, 26, 39, 52),
#     labels = c("Jan", "Apr", "Jul", "Oct", "Dec")
#   ) +
#   scale_y_continuous(limits = c(0, 100)) +
#   labs(
#     title    = "Norwegian reservoir fill rate: selected years vs. historical range",
#     subtitle = "Shaded band = historical min–max (1995–present). Dashed line = median.",
#     y        = "Fill rate (%)",
#     x        = NULL,
#     caption  = "Source: NVE Magasinstatistikk API"
#   ) +
#   theme_minimal(base_size = 11) +
#   theme(legend.position = "bottom")

# ── 7. GAP-FILLING (run only if has_gaps() returned TRUE in step 5c) ──────────
# NVE occasionally delays or skips a weekly publication if fewer than 90% of
# reservoir measurements have been received by Wednesday. This produces genuine
# missing rows in the tsibble index rather than NA values in existing rows.
# fill_gaps() inserts the missing index entries with NA values; the subsequent
# fill() then forward-fills those NAs from the most recent valid observation.
# Forward-filling is appropriate here because reservoir fill rate is a stock
# variable (it changes slowly and continuously), so the previous week's reading
# is the best available approximation for a missing week. If many consecutive
# weeks are missing, consider flagging those periods rather than forward-filling.

# fill_rate <- fill_rate %>%
#   fill_gaps() %>%                  # inserts NA rows for missing week indices
#   fill(fill_rate_pct,              # forward-fill from last valid observation
#        energy_GWh,
#        capacity_GWh,
#        .direction = "down")
```